---
title: "GETTING AND SUBSETTING DATA - COURSE PROJECT"
author: "David Sebastian Arcos Del Salto"
date: "2022-11-09"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_chunk$set(warning = FALSE)

library(rmarkdown)
library(knitr)

```

# Description

The `run_analysis.R` script performs the process for obtaining data and the 5 steps required For data cleaning and convertion into a tidy data set, as described in the course project’s instructions.

## 1. Load packages and get the data

Loading required packages

```{r, include=TRUE, echo=TRUE, eval=FALSE}
#install.packages(c("tidyverse", "data.table")) 
library(tidyverse) 
library(data.table)
```

For obtaining the data: 

1. Create a subdirectory named *"GetCleanProject"* to save the new data using the `file.exists()` and `dir.create()` functions.

2. Download the zip file from this [URL](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip) direction, and save it in the subdirectory created previously. 

3. Finally, using the `unzip()` command unzip the data included in the "UCI HAR Dataset" folder inside the zip file downloaded in the previous step and save it in the subdirectory.

```{r, include=TRUE, echo=TRUE, eval=FALSE}
if(!file.exists("./GetCleanProject")){dir.create("./GetCleanProject")}
download.file(URL, destfile = "./GetCleanProject/datafiles.zip", method = "curl")
unzip(zipfile = "./GetCleanProject/datafiles.zip", exdir = "./GetCleanProject")
```

## 2. Loading the data

### 2.1 Features

The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ.
```{r, include=TRUE, echo=TRUE, eval=FALSE}
features <- read.table("features.txt", 
                       col.names = c("N","Features")) # 561 obs x 1 variables
```

### 2.2 Activities

List of activities performed when the corresponding measurements were taken and its codes (labels included in Step.5).
```{r, include=TRUE, echo=TRUE, eval=FALSE}
activities <- read.table("activity_labels.txt", 
                         col.names = c("ID", "Activity")) # 6 obs x 2 variables
```

### 2.3 Train data

The train data contains data of 21/30 volunteer subjects being observed
```{r, include=TRUE, echo=TRUE, eval=FALSE}
subject_train <- read.table("train/subject_train.txt", 
                            col.names = "Volunteer") # 7.352 obs x 1 variable
y_train <- read.table("train/y_train.txt", 
                      col.names = "ActivityCode") # 7.352 obs x 1 variable
x_train <- read.table("train/X_train.txt", 
                      col.names = features$Features) # 7.352 obs x 561 variable
```

### 2.4 Test data

The test data contains data of 9/30 volunteer subjects being observed
```{r, include=TRUE, echo=TRUE, eval=FALSE}
subject_test <- read.table("test/subject_test.txt", 
                           col.names = "Volunteer") # 2.947 obs x 1 variable
y_test <- read.table("test/y_test.txt", 
                     col.names = "ActivityCode") # 2.947 obs x 1 variable
x_test <- read.table("test/X_test.txt", 
                     col.names = features$Features) # 2.947 obs x 561 variables
```

## 3. Merging the data

  This step merges the training and the test sets to create one data set, using the `rbind()` and `cbind()` functions: 

+ **Xset:** is created by merging the `x_train` and `x_test` sets.

+ **Yset:** is created by merging the `y_train` and `y_test` sets.

+ **Volunteer:** is created by merging the `subject_train` and `subject_test` sets.

+ **df:** is created by merging the three previous data sets into one big data set.

```{r, include=TRUE, echo=TRUE, eval=FALSE}
Xset <- rbind(x_train, x_test)
Yset <- rbind(y_train, y_test)
Volunteer <- rbind(subject_train, subject_test)
df <- cbind(Volunteer, Yset, Xset) # 10.299 obs x 563 variables
```

## 4. Extract the mean and standard deviation measurements

For extracting the mean and sd measurements, we use **regular expressions** to select the columns whose names contain *"mean"* or *"std"*. For instance, we use the expression `"(mean|std)\\(\\)` for selecting columns whose names include "mean" or "std", followed by a parenthesis "()". Also, we use the **"\\\\"** escape character to enable R to understand the parenthesis as part of the string. 

Further, we add **+2** to the index of the selected columns, so that the first and second column can be the identifier for *Volunteer* and *Activity*.  

```{r, include=TRUE, echo=TRUE, eval=FALSE}
columns_needed <- (grep("(mean|std)\\(\\)", features$Features)+2)
final_df <- df[, c(1,2,columns_needed)] # 10.299 obs x 68 variables
```

## 5. Set the appropriate labels for the activities and the variables

First, we match the **Activity code** with its correct label. In other words, the integers in the `ActivityCode` variable are replaced with the corresponding activity label from the *Activities* data set.

```{r, include=TRUE, echo=TRUE, eval=FALSE}
final_df$ActivityCode <- activities[final_df$ActivityCode, 2]
```

Secondly, we make sure the descriptive variable names are properly labeled. In essence, we rename the variable names so that they don't contain abbrevations, but instead contain the proper name for each variable. For this, we use the `gsub()` function, which identifies a string pattern inside a string and replaces it.   

```{r, include=TRUE, echo=TRUE, eval=FALSE}
names(final_df)[2] = "Activity" # Here, we change "ActivityCode" for just "Activity"
names(final_df)<-gsub("Acc", "Accelerometer", names(final_df))
names(final_df)<-gsub("Gyro", "Gyroscope", names(final_df))
names(final_df)<-gsub("BodyBody", "Body", names(final_df))
names(final_df)<-gsub("Mag", "Magnitude", names(final_df))
names(final_df)<-gsub("^t", "Time", names(final_df))
names(final_df)<-gsub("^f", "Frequency", names(final_df))
names(final_df)<-gsub("tBody", "TimeBody", names(final_df))
names(final_df)<-gsub("-mean()", "Mean", names(final_df), ignore.case = TRUE)
names(final_df)<-gsub("-std()", "STD", names(final_df), ignore.case = TRUE)
names(final_df)<-gsub("-freq()", "Frequency", names(final_df), ignore.case = TRUE)
names(final_df)<-gsub("angle", "Angle", names(final_df))
names(final_df)<-gsub("gravity", "Gravity", names(final_df))
```

## 6. Create an independent tidy data set

From the data set defined in step 4, we create an independent tidy data set with the mean of each variable, for each activity and for each volunteer. First, we create a data set that contains the mean for each variable using the `aggregate()` function. Then, we turn this data set into a tidy data set, using the `gather()` function. Finally, we export the tidy data set as a **.txt** file using the `write.table()` function.  

```{r, include=TRUE, echo=TRUE, eval=FALSE}
agg_df <- aggregate(. ~Volunteer + Activity, final_df, mean)

tidy_df <- gather(agg_df, Feature, Mean_value, -c(Volunteer, Activity)) %>% 
  arrange(Volunteer, Activity) 

write.table(tidy_df, file = "finaltidydata.txt",row.name=FALSE)

```